---
title: "Reproducible Research: Peer Assessment 1"
output: 
html_document:
keep_md: true
---


## Loading and preprocessing the data

We load some libraries that will be used in the this task. We also preprocess the activity data file, which is in the CSV format, removing NAs and formating columns. 
```{r ,echo = TRUE}
Sys.setlocale("LC_TIME", "C")
suppressWarnings(library("plyr",warn.conflicts = FALSE,quietly = TRUE))
suppressWarnings(library("lattice",warn.conflicts = FALSE,quietly = TRUE))
suppressWarnings(library("dplyr",warn.conflicts = FALSE,quietly = TRUE))
suppressWarnings(library("ggplot2",warn.conflicts = FALSE,quietly = TRUE))
setwd("C:\\Classes\\Coursera\\repdata-014")
activity<-read.csv("activity.csv",colClasses=c("integer", "Date", "integer"))
activity$interval<-60*floor((activity$interval+1)/100) + (activity$interval %% 100)
str(activity)
activity_h=na.omit(activity)
str(activity_h)
```

## What is mean total number of steps taken per day?

```{r }
total_steps <- tapply(activity_h$steps, activity_h$date, sum)
step_mean <- mean(total_steps)
sum_stepsxday=aggregate(x=list(steps=activity_h$steps),by=list(date=activity_h$date),sum)
mean_step=mean(sum_stepsxday$steps)
median_step=median(sum_stepsxday$steps)

```

The average number of steps per day is:
```{r}
step_mean
```

The histogram below shows the distribution of steps per day.
```{r}
hist(sum_stepsxday$steps,breaks=12,col = "green", xlab = "Steps", main="Histogram (steps x day)")
```

The mean and median are, respectively:
```{r}
mean_step
median_step
```


## What is the average daily activity pattern?

```{r}
avg_steps <- tapply(activity_h$steps, activity_h$interval, mean)
hours <- as.numeric(names(avg_steps))/60
plot(hours, avg_steps, type="l", axes=F,
     xlab="Time (h)", ylab="average steps in 5-min interval",
     main="Daily activity pattern")
     axis(1,at=0:24,labels=paste(0:24,":00", sep="")) 
     axis(2)
max_num<-which(avg_steps==max(avg_steps))
max_int<-activity_h$interval[max_num]

```

The 5-minute interval that contains the maximun number of steps is:
```{r}
sprintf("%02d:%02d", floor(max_int/60), max_int %% 60)
```

## Imputing missing values
Here we calculate the total number of NA values. We also try to fill in the missing values (5 min intervals). 
The histogram shows the total numbers of steps taken each day. Also the mean and median are calculated.   

```{r}
sum(is.na(activity))
activity_t <- transform(activity,steps=ifelse(is.na(steps),avg_steps, steps))
summary(activity_t)
steps_tot <- tapply(activity_t$steps, activity_t$date, sum, na.rm=T)
step_mean <- mean(steps_tot)
step_mean
step_median <- median(steps_tot)
step_median
hist(steps_tot, breaks=11, 
     xlab="steps per day", 
     sub="(with missing values)",
     main="Histogram of total steps x day")
abline(v=step_mean, col="blue", lwd=3)
abline(v=step_median, col="green", lwd=3, lty=2)
legend(x="topright", legend=c("mean","median"), col=c("blue","green"), bty="n", lwd=3)
sum(activity$steps, na.rm=TRUE)
sum(activity_t$steps)
```

## Are there differences in activity patterns between weekdays and weekends?

We create a factor variable with two levels, weekday and weekend. Then we plot time series data of the 5 min interval and the average number of steps taken comparating these two different approaches. 

```{r patterns}
week <- factor(weekdays(activity_t$date) %in% c("Saturday","Sunday"),labels=c("weekday","weekend"), ordered=FALSE)
tot_steps <- aggregate(activity_t$date, by=list(interval=activity_t$interval, weekday=week), mean)
activity_t$weekdays <- weekdays(as.Date(activity$date))
activity_t$weekdays <- ifelse(activity_t$weekdays %in% c("Saturday", "Sunday"),"weekend", "weekday")
average <- ddply(activity_t, .(interval, weekdays), summarise, steps=mean(steps))
xyplot(steps ~ interval | weekdays, data = average, layout = c(1, 2), type="l", xlab = "Interval", ylab = "# of steps")
```



